---
title: "Hands-On Exercise 4a"
author: "Albert Chia"
date: "May 07, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    page-layout: full
---

# Visualising Distribution

## Installing and loading the packages

In this exercise, we will be using the following R packages:

- ggridges - a ggplot2 extension specially designed for plotting ridgeline plot.
- ggdist - a ggplot2 extension spacially desgin for visualising distribution and uncertainty.
- tidyverse - a family of R packages to meet the modern data science and visual communication needs.
- ggthemes - a ggplot extension that provides the user additional themes, scales, and geoms for the ggplots package.
- colorspace - an R package provides a broad toolbox for selecting individual colors or color palettes, manipulating these colors, and employing them in various kinds of visualisations.

```{r}
pacman::p_load(ggdist, ggridges, ggthemes, colorspace, tidyverse)
```

## Importing Data

Similar to the previous Hands-on exercise, this exercise uses the *read_csv()* of readr package to import Exam_data.csv into R.

```{r}
#| code-fold: true
exam <- read_csv("Data/Exam_data.csv")
```

## Visualising Distribution with Ridgeline Plot

Ridgeline plot (aka Joyplot) is a data visualisation technique for revealing the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap.

There are several ways to plot ridgeline plot with R. In this exercise we will learn how to plot ridgeline plot using ggridges packages.

::: callout-note
## Note

ggridges package provides 2 main geom to plot gridgeline plots, they are: [geom_ridgeline()](https://wilkelab.org/ggridges/reference/geom_ridgeline.html) (takes hight values directly to draw the ridgelines) and [geom_density_ridges()](https://wilkelab.org/ggridges/reference/geom_density_ridges.html) (first estimates data densities and then draws those using ridgelines).
:::

The ridgeline plot below is plotted by using `geom_density_ridges()`

::: panel-tabset
## The Ridgeline Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()   
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()   
```
:::

::: callout-note
## Note

The key thing about this code is the new insert of geoom_density_ridges(). It includes the `scale` to control the height of the ridge; `rel_min_height` removes very tiny density tails, making the graph cleaner; and `bandwidth` controls the smoothness of the density curve.
:::

## Varying fill colors along the x axis

Sometimes we would like to have the area under a ridgeline not filled with a single solid color but rather with colors that vary in some form along the x axis. This effect can be achieved by using either [geom_ridgeline_gradient()](https://wilkelab.org/ggridges/reference/geom_ridgeline_gradient.html) or [geom_density_ridges_gradient()](https://wilkelab.org/ggridges/reference/geom_ridgeline_gradient.html). Both geoms work just like `geom_ridgeline()` and `geom_density_ridges()`, except that they allow for varying fill colors. However, they do not allow for alpha transparency in the fill. For technical reasons, we can have changing fill colors or transparency but not both.

::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()      
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()       
```
:::

::: callout-note
## Note

`geom_density_ridges_gradient` allows us to change the colour gradients. `viridis` is a colour palette package. It makes the plot nicer. `C` is just one of a viridis style.
:::

## Mapping the probabilities directly onto colour

Beside providing additional geom objects to support the need to plot ridgeline plot, ggridges package also provides a stat function called `stat_density_ridges()` that replaces `stat_density()` of ggplot2.

Figure below is plotted by mapping the probabilities calculated by using `stat(ecdf)` which represent the empirical cumulative density function for the distribution of English score.

:::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()            
```

## The Code

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()          
```

::: callout-note
## Note

`ecdf` = empirical cumulative distribution function. `fill = 0.5 - abs(0.5 - stat(ecdf))` - colour each part of the ridge based on how close it is to the middle of the distribution.
:::
::::

::: callout-important
## Important

It is important include the argument `calc_ecdf = TRUE` in `stat_density_ridges()`.
:::

## Ridgeline plots with quantile lines

By using `geom_density_ridges_gradient()`, we can colour the ridgeline plot by quantile, via the calculated `stat(quantile)` aesthetic as shown in the figure below.

:::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()           
```

## The Code

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()          
```

::: callout-note
## Note

- Instead of colouring continuously across the x-axis, it divides the distribution into 4 sections (quartiles).
- `scale_fill_viridis_d` - discrete because quartiles are categories.
:::
::::

Instead of using number to define the quantiles, we can also specify quantiles by cut points such as 2.5% and 97.5% tails to colour the ridgeline plot as shown in the figure below.

::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()             
```

## The Code

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()            
```
:::

## Visualising Distribution with Raincloud Plot

Raincloud Plot is a data visualisation techniques that produces a half-density to a distribution plot. It gets the name because the density plot is in the shape of a “raincloud”. The raincloud (half-density) plot enhances the traditional box-plot by highlighting multiple modalities (an indicator that groups may exist). The boxplot does not show where densities are clustered, but the raincloud plot does!

In this section, you will learn how to create a raincloud plot to visualise the distribution of English score by race. It will be created by using functions provided by **ggdist** and ggplot2 packages.

## Plotting a Half Eye graph

First, we will plot a Half-Eye graph by using `stat_halfeye()` of ggdist package.

This produces a Half Eye visualization, which is contains a half-density and a slab-interval.

:::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)            
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)           
```

::: callout-note
## Things to learn from the code chunk above

We remove the slab interval by setting `.width = 0` and `point_colour` = NA.
:::
::::

## Adding the boxplot with 'geom_boxplot()'

Next, we will add the second geometry layer using [geom_boxplot()](https://r4va.netlify.app/chap09) of ggplot2. This produces a narrow boxplot. We reduce the width and adjust the opacity.

::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)          
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)        
```
:::

## Adding the Dot Plots with `stat_dots()`

Next, we will add the third geometry layer using [stat_dots()](https://mjskay.github.io/ggdist/reference/stat_dots.html) of ggdist package. This produces a half-dotplot, which is similar to a histogram that indicates the number of samples (number of dots) in each bin. We select side = “left” to indicate we want it on the left-hand side.

:::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)            
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)            
```

::: callout-note
## Note

- `adjust = 0.5` - Controls smoothness of the density curve. Smaller means more wavy; larger means smoother.
:::
::::

## Finishing Touch

Lastly, [coord_flip()](https://ggplot2.tidyverse.org/reference/coord_flip.html) of ggplot2 package will be used to flip the raincloud chart horizontally to give it the raincloud appearance. At the same time, `theme_economist()` of ggthemes package is used to give the raincloud chart a professional publishing standard look.

::: panel-tabset
## The Plot

```{r echo=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()           
```

## The Code

```{r eval=FALSE}
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()           
```
:::
