---
title: "Hands-on_Ex03a - Programming Interactive Data Visualisation with R"
author: "Albert Chia"
date: "April 30, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: false
---

# Programming Interactive Data Visualisation with R

## Installing and loading the required libraries

In the Hands-on exercise, we will learn to create interactive data visualisation by using more functions provided by ggiraph and plotlyr packages. First, we should install and launch the following R packages:

- ggiraph for making ‘ggplot’ graphics interactive.
- plotly, R library for plotting interactive statistical graphs.
- DT provides an R interface to the JavaScript library DataTables that create interactive table on html page.
- tidyverse, a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.
- patchwork for combining multiple ggplot2 graphs into one figure.

```{r}
pacman::p_load(ggiraph, plotly, patchwork, DT, tidyverse)
```

## Importing Data

Similar to the previous Hands-on exercise, this exercise uses the read_csv() of readr package to import Exam_data.csv into R.

```{r}
exam_data <- read_csv("Data/Exam_data.csv")
```

## Interactive Data Visualisation - ggiraph methods

ggiraph is an htmlwidget and a ggplot2 extension. It allows ggplot graphics to be interactive.

Interactive is made with ggplot geometries that can understand 3 arguments:

1.  Tooltip: a column of data-sets that contain tooltips to be displayed when the mouse is over elements.
2.  Onclick: a column of data-sets that contain a JavaScript function to be executed when elements are clicked.
3.  Data_id: a column of data-sets that contain an id to be associated with elements.

If used within a shiny application, elements associated with an id(data_id) can be selected and manipulated on clinet and server sides.

## All about Tooltip

:::: panel-tabset
## Tooltip Effect with Tooltip Asethetic

Below shows an interactive statistical plot using the ggiraph package.

```{r}
#| code-fold: true

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

Notice that two steps are involved. First, an interactive version of ggplot2 geom (i.e. geom_dotplot_interactive()) will be used to create the basic graph. Then, girafe() will be used to generate an svg object to be displayed on an html page.

::: callout-note
## Note

- "#\| code-fold: true" hide the code chunk, and reveals upon click.
- aes(tooltip=id) enables the hovering over a dot to show 'id' (This is from ggiraph) stackgroups = TRUE basically just stacks the dots nicely without overlapping.
- girafe() converts the ggplot into a interactive HTML widget ggobj = p means to use the plot 'p' we assigned in the first line.
:::

## Displaying multiple information on tooltip

The content of the tooltip can be customised by including a list object as shown. The first 3 lines of codes in the code chunk create a new field called tooltip. At the same time, it populates text in ID and CLASS fields into the newly created field. Next, this newly created field is used as tooltip field as shown in the code of line 7.

```{r}
#| code-fold: true

exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```

::: callout-note
## Note

- "exam_data\$tooltip \<-" creates a column called 'tooltip'.
- aes(tooltip = exam_data\$tooltip) basically means that now tooltip shows multi-line info instead of just 'id'.
:::

## Customising Tooltip style

Below uses opts_tooltip() of ggiraph to customise tooltip rendering by adding css declarations. Notice how the css changed the background colour of the tooltip to black and the font colour to white and bold.

```{r}
#| code-fold: true

tooltip_css <- "background-color:white; font-weight:bold; color:black;" 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = ID),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(opts_tooltip(css = tooltip_css))
)
```

::: callout-note
## Note

- "tooltip_css \<-" defines the style i want; "options = list(opts_tooltip(css = tooltip_css)) is for ggiraph to apply in when rendering. font-weight changes bold; font-style changes italic.
:::

## Displaying statistics on tooltip

This shows an advanced way to customise tooltip. The example below is used to compute 90% confident interval of the mean. The derived statistics are then displayed in the tooltip.

```{r}
#| code-fold: true

tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

::: callout-note
## Note

- The first line defines a function - one that adds the mean and the standard error (SEM).
- "paste" just means that the text is added to the chart.
- "stat_summary" calculates the stats.
- "after_stats()" uses those calculated values. (y = mean, ymax = mean + SEM)
- "fun.data = "mean_se" simply computes the mean and standard error.
- "geom = GeomInteractiveCol" draws the bars (column) and makes it interactive (meaning the hovering works).
- Everything in the code did 3 key things: (1) Compute summary stats, (2) Plots them, and (3) Add smart tooltip.
:::
::::

## Hover

:::: panel-tabset
## Hover Effect with data_id Aesthetic

This is the 2nd interactive feature of ggiraph, namely data_id. Elements associated with a data_id (i.e.CLASS) will be highlighted upon mouse over.

```{r}
#| code-fold: true

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(           
    aes(tooltip = ID, 
        data_id = CLASS),             
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)    
```

::: callout-note
## Note

- One obvious changes is the above uses data_id and not tooltip.
- i added "aes(tooltip = ID)" so that hovering now shows the ID as well. Just for fun.
- the interactive shows dots from the same CLASS being highlighted together.
:::

## Styling hover effect

In the code chunk below, css codes are used to change the highlighting effect.

```{r}
#| code-fold: true

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)   
```

::: callout-note
## Note

- Options now list the hover behaviour - filling the hovered dots and fading the rest of the dots.
- But this plot did not display the CLASS of the selected dot (which i will show next).
:::

## Combining Tooltip and Hover Effect

There are time that we want to combine tooltip and hover effect on the interactive statistical graph. Interactivity: Elements associated with a data_id (i.e CLASS) will be highlighted upon mouse over. At the same time, the tooltip will show the CLASS.

```{r}
#| code-fold: true

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)     
```

::: callout-note
## Note

- Basically, this is the same as what i did for fun with the first Hover Effect - adding the aes(tooltip = CLASS). This adds the effect when you hover, it shows the CLASS.
- data_id = CLASS linked all the dots with the same CLASS.
:::

## Click Effect with onclick

onclick argument of ggiraph provides hotlink interactivity on the web. Interactivity: Web document link with a data object will be displayed on the web browser upon mouse click.

```{r}
#| code-fold: true

exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618) 
```

::: callout-note
## Note

- Similiarly, we created a new column called onclick.Each row contains a javacript.
- "sprintf() creates a string based on (%s%s, URL, ID), which basically adds the url to an ID behind.
- aes (onclick = onclick) tells R that whenever i click a dot, run the javascript stored in that row.
:::

## Coordinated Multiple Views with ggiraph

- Coordinated multiple views methods has been implemented in the data visualisation below.

- Notice that when a data point of one of the dotplot is selected, the corresponding data point ID on the second data visualisation will be highlighted too.

- In order to build a coordinated multiple views as shown in the example above, the following programming strategy will be used:

  1.  Appropriate interactive functions of ggiraph will be used to create the multiple views.
  2.  Patchwork function of patchwork package will be used inside girafe function to create the interactive coordinated multiple views.

```{r}
#| code-fold: true

p1 <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       )  
```

::: callout-note
## Note
:::
- Basically, this enable us to generate a linked interactive visualisation - having 2 plots that "talks" to each other using data_id.

- Above shows the maths stats with english stats side-by-side.

- P1 and P2 are linked by the unique ID via "data_is = ID".

- "girafe(code = print(p1+p2))" tell R to put plots side-by-side, and makes them interactive together.
::::

## Interactive Data Visualisation - plotly methods!

Plotly’s R graphing library create interactive web graphics from ggplot2 graphs and/or a custom interface to the (MIT-licensed) JavaScript library plotly.js inspired by the grammar of graphics. Different from other plotly platform, plot.R is free and open source.

There are two ways to create interactive graph by using plotly, they are:

- by using plot_ly(), and
- by using ggplotly()

## Creating an Interactive Scatter Plot: plot_ly method
The tabset below shows an example a basic interactive plot created by using plot_ly(). 

:::: panel-tabset 
## The Plot
```{r echo=FALSE}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```
## The Code
```{r eval=FALSE}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```
::::

## Working with VIsual Variable: plot_ly method

Below is a example of colour argument mapped to a qualitative visual variable (i.e.RACE). 

:::: panel-tabset 
## The Plot

```{r echo=FALSE}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

## The Code

```{r eval=FALSE}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```
::::

## Coordinated Multiple Views with plotly

The creation of a coordinated linked plot by using plotly involves three steps:

- Highlight_key() of plotly package is used as shared data.
- Two scatterplots will be created by using ggplot2 functions.
- Lastly, subplot() of plotly package is used to place them next to each other side-by-side.

Thing to learn from the code chunk:

- Highlight_key() simply creates an object of class crosstalk::SharedData.
- Visit this [link](https://rstudio.github.io/crosstalk/) to learn more about crosstalk.

:::: panel-tabset
## The Plot

```{r echo=FALSE}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```

## The Code Chunk

```{r eval=FALSE}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```
::::
::: callout-note
## Note
- plotly + crosstalk linking is another way to make plots interact with each other, beside ggiraph.
- The main idea is to create a shared data. This is case, it is the "d \<- highlight_key(exam_data)"
- p1 and p2 uses "data=d", which linked them together.
- ggplotly" converts them into interactive plotly.
:::

## Interactive Data Visualisation - crosstalk methods!

[crosstalk](https://rstudio.github.io/crosstalk/index.html) is an add-on to the htmlwidgets package. It extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions (currently, linked brushing and filtering).

## Interactive Data Table: DT Package

- A wrapper of the JavaScript Library DataTables.
- Data objects in R can be rendered as HTML tables using the JavaScript library 'DataTables' (typically via R Markdown and Shiny).

```{r}
#| code-fold: true

DT::datatable(exam_data, class= "compact")
```

## Linked Brushing: crosstalk method

Things to learn:

- highlight() is a function of plotly package. It sets a variety of options for brushing (i.e., highlighting) multiple plots. These options are primarily designed for linking multiple plotly graphs, and may not behave as expected when linking plotly to another htmlwidget package via crosstalk. In some cases, other htmlwidgets will respect these options, such as persistent selection in leaflet.

- bscols() is a helper function of crosstalk package. It makes it easy to put HTML elements side by side. It can be called directly from the console but is especially designed to work in an R Markdown document. Warning: This will bring in all of Bootstrap!

:::: panel-tabset
## The Plot

```{r echo=FALSE}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```

## The Code Chunk

```{r eval=FALSE}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```
::::
